How to \emph{not} write Virtual Machines for Dynamic Languages

### 2007-07-15 talk/dyla2007/dyla.tex
Some more progress.

Dynamic languages are traditionally implemented by writing a virtual
machine for them in a low-level language like C or in a language that
can relatively easily be turned into C.  The machine implements an
object model supporting the high level dynamic language's objects.  It
typically provides features like automatic garbage collection.  Recent
languages like Python, Ruby, Perl and JavaScript have complicated
semantics which are most easily mapped to a naive interpreter operating
on syntax trees or bytecode; simpler languages\footnote
{
In the sense of the primitive semantics.  ``Simple'' here is
as opposed to ``complicated'', not as opposed to ``complex'': Common
Lisp for example is not a small language, but it can at least in theory
be expressed from a smaller core of primitives.  In Python, all
primitive operations have complicated semantics.  The argument developed
in the present paper is more relevant to ``complicated'' dynamic languages.
}
like Lisp, Smalltalk and Self typically have more
efficient implementations based on code generation.

The effort required to build a new virtual machine is relatively
large.  This is particularly true for languages which are complex
and in constant evolution. Language implementation communities from an
open-source or academic context have only limited resources. Therefore they
cannot afford to have a highly complex implementation and often choose simpler
techniques even if that entails lower execution speed. Similarly, fragmentation
(for example because of other implementations of the same language) is a
problem because it divides available resources. All these points also apply to
the implementation of domain-specific languages where it is important to keep
the implementation effort small.

For these reasons writing a virtual machine in C is problematic because it
forces the language implementer to deal with many low-level details (like
garbage collection and threading issues). Limitations
of the C implementation lead to alternative implementations which draw
resources from the reference implementation. An alternative to writing
implementations in C is to build them on top of one of the newer object oriented
virtual machines (``OO VM'') such as the JVM (Java Virtual Machine) or the
CLR (Common Language Runtime of the .NET framework). This is often wanted by
the community anyway, since it leads to the ability to re-use the libraries of
these platforms. However, if a C implementation existed before the
implementation of such a VM is started, this enters into conflict with the goal of
having to maintain essentially a single, simple enough implementation for a
given programming language: as the language becomes popular, there will be a
demand to have it run on various platforms -- high-level VMs as well as
C-level environments.

In this paper, we will argue that it is possible to
benefit from and integrate with OO VMs while keeping the dynamic
language implemented with a single, simple source code base.  The idea is
to write an interpreter for that language in another sufficiently
high-level but less dynamic language.  This interpreter plays the role
of a specification for the dynamic language.  With a sufficiently capable
translation toolchain we can then generate whole virtual machines from
this specification -- either wholly custom VMs for C-level operating
systems or as a layer on top of various OO VMs.  In other words,
meta-programming techniques can be used to successfully replace a
foreseeable one-VM-fits-all standardization attempt.

@@

PyPy contains a Python interpreter implemented in Python, from which
Python VMs can be generated.  The reader is referred to
\cite{pypyvmconstruction} for a technical presentation.  Let us emphasis
again that the argument we make in the present paper is \emph{not} that
VMs for dynamic languages should be written in their own host language
(as many projects like Squeak \cite{Squeak} and others have done) but
instead that VMs should not be \emph{written} in the first place -- they
should be generated from simple interpreters written in any suitable
high-level\footnote{``High-level'' is taken by opposition to languages
like Scheme48's PreScheme \cite{kelsey-prescheme} or Squeak's \cite{Squeak}
SLang which use the syntax and
metaprogramming facilities of a high-level language but encode
low-level details like object layout and memory management.} language.

In section \ref{sect:approaches} we will explore the way VMs are typically
implemented in C and on top of OO VMs and some of the problems of these
approaches, using various Python implementations as the main example. In
section \ref{sect:metaprogramming} we will describe our proposed
meta-programming approach.

@@

The observation that limitations of a C-based implementation of a dynamic
language leads to the emergence of additional implementations is clear
in the case of Python.  The reference implementation, \implname{CPython}
\cite{cpy251}, is a simple recursive interpreter.  \implname{Stackless
Python} \cite{stackless} is a fork that adds micro-threading
capabilities to CPython. One of the reasons for not incorporating it back
into CPython was that it was felt that this would make the
implementation too complex. Another implementation of the Python
language is \implname{Psyco} \cite{psyco-software}, an extension of
CPython which adds a JIT-compiler.  Finally, \implname{Jython} is a
re-implementation for the Java VM and \implname{IronPython} for
the CLR.  All of these ultimately need to be kept synchronized with the
relatively fast evolution of the language.

@@

Implementing a language on top of an existing OO VM is in many ways easier than
implementing it in C. Let's take a look at the advantages that are usually
cited for basing a
language implementation of a dynamic language on a standard object oriented
virtual machine, for example the JVM or the CLR.

@@

\section{Conclusion}

>>>>

Dynamic languages are traditionally implemented by writing a virtual
machine (VM) for them in a low-level language like C or in a language that
can relatively easily be turned into C.  The VM implements an
object model supporting the high level dynamic language's objects.  It
typically provides features like automatic garbage collection.  Recent
languages like Python, Ruby, Perl and JavaScript have complicated
semantics which are most easily mapped to a naive interpreter operating
on syntax trees or bytecode; simpler languages\footnote
{
In the sense of the primitive semantics.  For example, in Python most
primitive operations have complicated semantics; by contrast, in Common
Lisp complex features like the reader and printer can in theory be
implemented in terms of simpler primitives as library code.
}
like Lisp, Smalltalk and Self typically have more
efficient implementations based on code generation.

The effort required to build a new virtual machine is relatively
large.  This is particularly true for languages which are complicated
and in constant evolution. Language implementation communities from an
open-source or academic context have only limited resources. Therefore they
cannot afford to have a highly complex implementation and often choose simpler
techniques even if that entails lower execution speed. Similarly, fragmentation
(for example because of other implementations of the same language) is a
problem because it divides available resources. All these points also apply to
the implementation of domain-specific languages where it is important to keep
the implementation effort small.

For these reasons writing a virtual machine in C is problematic because it
forces the language implementer to deal with many low-level details (like
garbage collection and threading issues). If a language becomes popular,
limitations of the C implementation eventually
lead to alternative implementations which draw
resources from the reference implementation. An alternative to writing
implementations in C is to build them on top of one of the newer
general-purpose object-oriented
virtual machines (``OO VM'') such as the JVM (Java Virtual Machine) or the
CLR (Common Language Runtime of the .NET framework). This is often wanted by
the community anyway, since it leads to the ability to re-use the libraries of
these platforms. However, if a C implementation existed before the
implementation of such a VM is started, this enters into conflict with the goal of
having to maintain essentially a single, simple enough implementation for a
given programming language: as the language becomes popular, there will be a
demand to have it run on various platforms -- high-level VMs as well as
C-level environments.

In this paper, we will argue that it is possible to
benefit from and integrate with OO VMs while keeping the dynamic
language implemented as a single, simple source code base.  The idea is
to write an interpreter for that language in another sufficiently
high-level but less dynamic language.  This interpreter plays the role
of a specification for the dynamic language.  With a sufficiently capable
translation toolchain we can then generate whole virtual machines from
this specification -- either wholly custom VMs for C-level operating
systems or as a layer on top of various OO VMs.  In other words,
meta-programming techniques can be used to successfully replace a
foreseeable one-OO-VM-fits-all standardization attempt.

@@

PyPy contains a Python interpreter implemented in Python, from which
Python VMs can be generated.  The reader is referred to
\cite{pypyvmconstruction} for a technical presentation.  Let us emphasis
that the argument we make in the present paper is \emph{not} that
VMs for dynamic languages should be written in their own host language
(as many projects like Squeak \cite{Squeak} and others have done) but
instead that VMs should not be \emph{written} in the first place -- they
should be generated from simple interpreters written in any suitable
high-level\footnote{``High-level'' is taken by opposition to languages
like Scheme48's PreScheme \cite{kelsey-prescheme} or Squeak's \cite{Squeak}
SLang, which use the syntax and
metaprogramming facilities of a high-level language but encode
low-level details like object layout and memory management.} language.

In section \ref{sect:approaches} we will explore the way VMs are typically
implemented in C and on top of OO VMs and some of the problems of these
approaches, using various Python implementations as the main example. In
section \ref{sect:metaprogramming} we will describe our proposed
meta-programming approach and compare the two solutions.  We summarize
our position and conclude in section \ref{sect:conclusion}.

@@

Limitations of a C-based implementation of a dynamic
language lead to the emergence of additional implementations --
this observation is clear
in the case of Python.  The reference implementation, \implname{CPython}
\cite{cpy251}, is a simple recursive interpreter.  \implname{Stackless
Python} \cite{stackless} is a fork that adds micro-threading
capabilities to CPython. One of the reasons for not incorporating it back
into CPython was that it was felt that this would make the
implementation too complex. Another implementation of the Python
language is \implname{Psyco} \cite{psyco-software}, an extension of
CPython which adds a JIT-compiler.  Finally, \implname{Jython} is a
re-implementation for the Java VM and \implname{IronPython}
a re-implementation for
the CLR.  All of these ultimately need to be kept synchronized with the
relatively fast evolution of the language.

@@

Implementing a language on top of an existing OO VM is in many ways easier than
implementing it in C. Let's take a look at the advantages that are usually
cited for basing a
language implementation of a dynamic language on a standard object oriented
virtual machine.

@@

\section{Conclusion}
\label{sect:conclusion}
